---
title: "Monte-Carlo Consistent Partial Least Squares"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Monte-Carlo Consistent Partial Least Squares}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(plssem)
```

In the `plssem()` package we've implemented a version of the
*Monte-Carlo Consistent Partial Least Squares* algorithm (MC-PLSc).
Here we present a general explanation of the algorithm. A more detailed
explanation can be found [here](https://osf.io/preprints/psyarxiv/fwzj6_v1).

The general idea of the MC-PLSc algorithm is to treat the PLS algorithm as a
biased estimator under a given set of circumstances. The PLS algorithm is
in general a biased estimator of common factor models (i.e., reflective constructs),
where additional biased is introduced when the PLS algorithm is applied to
ordinal data, as if the data is continuous.

The MC-PLSc algorithm then uses a Monte-Carlo procedure to model the bias of
the PLS algorithm under the given circumstance, solving for a consistent set
of model parameters.

# Example

Consider the correlation $r$ between continous variables $x$ and $y$. 

```{r}
set.seed(2348929)
n <- 10000
r <- 0.6
x <- rnorm(n)
y <- r * x + rnorm(n, sd = sqrt(1 - r^2))
cor(x, y)
```

Let's then say that we only have access to a set of binary representations of
$x$ and $y$, $z$ and $w$. Here we just split $x$ and $y$ in the middle.

```{r}
z <- as.integer(x > 0)
w <- as.integer(y > 0)
```

If we then calculate the correlation between $z$ and $w$ we will get a biased
estimate of $r$.

```{r}
cor(z, w)
```

While the estimate is biased, it does say something about the true correlation.
The goal of the MC-PLSc algorithm is to model how this bias behaves. Let's
write a function `g()` which simulates data from our model, given an estimate
of $r$, and a function `f()` for our estimator, which in this case is the correlation
coefficient.

```{r}
g <- function(r_hat, R = 100000) {
  # Generate continuous data
  x <- rnorm(R)
  y <- r_hat * x + rnorm(R, sd = sqrt(1 - r_hat^2))

  # Generate binary data
  z <- as.integer(x > 0)
  w <- as.integer(y > 0)

  # Return a data.frame
  data.frame(z, w)
}

f <- function(df) {
  cor(df$z, df$w)
}
```

Putting `f()` and `g()` together we can mess around with different values of 
`r_hat` to find an estimate which matches the observed correlation of $z$ and $w$

```{r}
print(f(g(0.0)))
print(f(g(0.3)))
print(f(g(0.5)))
print(f(g(0.6)))
```

Here we can see that our original value for $r$ (`0.6`) get's us quite close. However
we can do better than manual trial an error. Therefore we rephrase the problem as a
root finding problem.

```{r}
r_biased_observed <- cor(z, w)
h <- function(r_hat) {
  f(g(r_hat)) - r_biased_observed
}
```

Using the biased estimate as our estimate we can run a stochastic root approximation
using the `SimDesign` package.

```{r}
r_hat <- r_biased_observed # good starting estimate
iter <- 10
history <- r_hat

for (i in seq_len(iter)) {
  temperature <- 1 / sqrt(i)
  epsilon <- h(r_hat)
  r_hat <- r_hat - temperature * epsilon
  history <- c(history, r_hat)
}

plot(history)
```

```{r}
library(ggplot2)
data.frame(y = history, x = seq_along(history)) |>
ggplot(aes(x = x, y = y)) + geom_line() + geom_point() + geom_smooth(se = FALSE)
```
